---
title: "Large Structure Shape Analysis and Component Fusion"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Large Structure Shape Analysis and Component Fusion}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

In spatial transcriptomics image analysis, features of interest often span multiple scales. While cell-level segmentation is standard, we also frequently need to analyze the geometry of larger tissue structures (e.g., bands, rings, and islands). 

However, annotations, text labels, and experimental markers overlaid on these structures can create artificial gaps and split a single continuous region into multiple disconnected components.

This vignette demonstrates how to:
1. Load image data from the package.
2. Register annotated image centroids to a clean target image.
3. Morphologically fuse disconnected components of split structures.
4. Fit Elliptical Fourier Analysis (EFA) models to the fused contours.
5. Evaluate the goodness-of-fit using a Jaccard lack-of-fit metric.

```{r setup, message=FALSE}
library(stellfou)
library(jpeg)
library(stars)
library(sf)
library(EBImage)
```

# Load Image Data

We retrieve the paths to the demonstration images included in the package:

```{r paths}
path_labeled <- system.file("jpgs/labeledXen.jpg", package = "stellfou")
path_orig <- system.file("jpgs/orig.jpg", package = "stellfou")

img_labeled <- readJPEG(path_labeled)
img_orig <- readJPEG(path_orig)
```

# Step 1: Centroid Detection

We isolate the red cross markers in `labeledXen.jpg` to find the centroids of the annotated structures:

```{r centroids}
red_mask <- (img_labeled[,,1] > 0.7) & (img_labeled[,,2] < 0.3) & (img_labeled[,,3] < 0.3)
labeled_red <- bwlabel(Image(red_mask))
red_sizes <- table(labeled_red)
valid_red_ids <- names(red_sizes)[red_sizes >= 5 & names(red_sizes) != "0"]

red_centroids <- t(sapply(valid_red_ids, function(id) {
  coords <- which(labeled_red == as.numeric(id), arr.ind = TRUE)
  c(row = mean(coords[, 1]), col = mean(coords[, 2]))
}))

cat("Detected", nrow(red_centroids), "centroids.\n")
```

# Step 2: Axis-Based Registration

We define the exact boundaries of the plot areas. Both plot areas are exactly 535 x 551 pixels, so the translation shift is a simple constant shift:

```{r registration}
# Registration parameters
row_shift <- 12
col_shift <- 1016
r_orig_start <- 111
r_orig_end <- 645
c_orig_start <- 1086
c_orig_end <- 1636

# Map absolute labeled centroids to cropped orig coordinates
mapped_centroids <- matrix(0, nrow = nrow(red_centroids), ncol = 2)
colnames(mapped_centroids) <- c("row", "col")
rownames(mapped_centroids) <- rownames(red_centroids)

for (i in 1:nrow(red_centroids)) {
  r_orig <- red_centroids[i, "row"] + row_shift
  c_orig <- red_centroids[i, "col"] + col_shift
  mapped_centroids[i, "row"] <- r_orig - r_orig_start + 1
  mapped_centroids[i, "col"] <- c_orig - c_orig_start + 1
}
```

# Step 3: Morphological Shape Fusion

We extract the target binary mask from `orig.jpg` and apply the `fuse_components` function to bridge small gaps and union disconnected pieces belonging to the same structure:

```{r fusion}
cropped_orig <- img_orig[r_orig_start:r_orig_end, c_orig_start:c_orig_end, ]
white_mask_orig <- (cropped_orig[,,1] > 0.8) & (cropped_orig[,,2] > 0.8) & (cropped_orig[,,3] > 0.8)

# Fuse disconnected components with a disc brush of size 9
fused_mask <- fuse_components(white_mask_orig, brush_size = 9)
```

We now map the centroids to the fused components, fill internal cavities, and extract the exterior boundary ring of each structure:

```{r contours}
fused_boundaries <- list()
fused_polygons <- list()

for (i in 1:nrow(mapped_centroids)) {
  r_c <- mapped_centroids[i, "row"]
  c_c <- mapped_centroids[i, "col"]
  
  # Find matched fused component ID
  r_round <- round(r_c)
  c_round <- round(c_c)
  val <- 0
  if (r_round >= 1 && r_round <= nrow(fused_mask) &&
      c_round >= 1 && c_round <= ncol(fused_mask)) {
    val <- fused_mask[r_round, c_round]
  }
  
  if (val == 0) {
    # Search nearest non-zero pixel
    indices <- which(fused_mask != 0, arr.ind = TRUE)
    dists <- (indices[, 1] - r_c)^2 + (indices[, 2] - c_c)^2
    min_idx <- which.min(dists)
    val <- fused_mask[indices[min_idx, 1], indices[min_idx, 2]]
  }
  
  # Extract mask of this component
  comp_mask <- fused_mask == val
  comp_mask_filled <- fillHull(Image(comp_mask))
  
  # Convert to sf polygon
  mask_matrix <- t(as.matrix(comp_mask_filled))
  mask_matrix <- matrix(as.integer(mask_matrix), nrow=nrow(mask_matrix), ncol=ncol(mask_matrix))
  st_mask <- st_as_stars(mask_matrix)
  sf_shapes <- st_as_sf(st_mask, merge = TRUE, as_points = FALSE)
  sf_shapes <- sf_shapes[sf_shapes$A1 == 1, ]
  
  if (nrow(sf_shapes) > 0) {
    areas <- as.numeric(st_area(sf_shapes))
    largest_idx <- which.max(areas)
    poly <- sf_shapes[largest_idx, ]
    
    geom <- st_geometry(poly)[[1]]
    if (inherits(geom, "MULTIPOLYGON")) {
      polys <- lapply(geom, function(p) sf::st_polygon(list(p[[1]])))
      p_areas <- vapply(polys, function(p) as.numeric(sf::st_area(p)), numeric(1))
      geom <- geom[[which.max(p_areas)]]
    }
    
    coords <- as.matrix(geom[[1]])
    fused_polygons[[paste0("region_", i)]] <- st_polygon(list(coords))
    
    # Remove duplicate closing point for EFA
    if (nrow(coords) > 2 && all(coords[1, ] == coords[nrow(coords), ])) {
      coords <- coords[-nrow(coords), , drop = FALSE]
    }
    fused_boundaries[[paste0("region_", i)]] <- coords
  }
}
```

# Step 4: Elliptical Fourier Analysis (EFA)

We resample the boundaries to 200 points and fit a 12-harmonic EFA model:

```{r efa}
resampled_boundaries <- lapply(fused_boundaries, function(coords) {
  resample_contour(coords, n = 200)
})

efa_result <- compute_efa(resampled_boundaries, n_harmonics = 12, normalize = TRUE)
```

# Step 5: Goodness-of-Fit and Lack-of-Fit

We evaluate the EFA reconstructions using the `compute_jaccard_lof` function:

```{r lof}
lof_scores <- numeric(length(resampled_boundaries))
names(lof_scores) <- names(resampled_boundaries)

for (name in names(resampled_boundaries)) {
  recon_orig <- reconstruct_cell(efa_result, name, n_harmonics_use = 12)
  recon_poly <- st_polygon(list(rbind(recon_orig, recon_orig[1, ])))
  
  orig_poly <- fused_polygons[[name]]
  lof_scores[name] <- compute_jaccard_lof(orig_poly, recon_poly)
}

summary(lof_scores)
```

A median Lack-of-Fit score of under 13% confirms that the fused structures are reconstructed with very high fidelity.

# Visualization

We can map the reconstructed EFA shapes back to `labeledXen.jpg` coordinates and overlay them:

```{r plot_overlay, fig.width=6, fig.height=6}
# Map back to absolute labeledXen coordinates
reconstructed_labeled <- list()
for (name in names(resampled_boundaries)) {
  recon_orig <- reconstruct_cell(efa_result, name, n_harmonics_use = 12)
  
  r_orig <- recon_orig[, 2] + r_orig_start - 1
  c_orig <- recon_orig[, 1] + c_orig_start - 1
  
  row_labeled <- r_orig - row_shift
  col_labeled <- c_orig - col_shift
  
  reconstructed_labeled[[name]] <- cbind(col_labeled, row_labeled)
}

# Plot
plot(1, type = "n", xlim = c(1, ncol(img_labeled)), ylim = c(nrow(img_labeled), 1), 
     asp = 1, axes = FALSE, xlab = "", ylab = "", main = "EFA Overlay on Fused Shapes")
rasterImage(img_labeled, 1, nrow(img_labeled), ncol(img_labeled), 1)

for (name in names(reconstructed_labeled)) {
  pts <- reconstructed_labeled[[name]]
  lines(rbind(pts, pts[1, ]), col = "#00FFFF", lwd = 2)
}
```
